---
title: "Tooth Growth Analysis"
output: pdf_document
---

---
```{r Load libraries, results="hide", echo=FALSE}
suppressWarnings(library(dplyr, quietly=TRUE, warn.conflict=FALSE))
library(scales)
library(datasets)
```

Emploration of the R ToothGrowth data set, which tracks the effect of vitamin C on tooth growth in guinea pigs. The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods: orange juice (**OJ**), or ascorbic acid (**VC**).

### 1. Basic exploratory data analysis 
```{r echo=TRUE}
data("ToothGrowth")
head(ToothGrowth)
```

We have 10 samples per dose for each of the two supplements.  Below is a histogram that gives an idea of the spread in tooth length for each of the 6 supplement:dose combinations.  As we can see the number of observations is small, and it is not really clear whether they have a normal distribution.  **Assumption:** the data has a normal distribution. 

```{r echo=FALSE}
par(mar=c(4,4,2,2))
par(mfrow=c(2, 3))
for(s in c("VC", "OJ")) for(d in c(0.5, 1.0, 2.0)) hist(ToothGrowth[ToothGrowth$supp == s & ToothGrowth$dose == d, ]$len, xlab = paste(s, ":", d), main = NULL, breaks=10)
```

### 2. Basic summary of the data.

```{r}
tg_oj <- subset(ToothGrowth, supp == "OJ")
tg_vc <- subset(ToothGrowth, supp == "VC")
tg_vc$difference = tg_vc$len - tg_oj$len
tg_oj$difference = tg_oj$len - tg_vc$len

n <- 10
tg_oj <- summarise(group_by(tg_oj, Supplement = supp, Dose = dose), 
                   MeanToothLength = mean(len), MeanDiff = mean(difference), 
                   SDDiff = sd(difference), N=length(len))
tg_vc <- summarise(group_by(tg_vc, Supplement = supp, Dose = dose), 
                   MeanToothLength = mean(len), MeanDiff = mean(difference), 
                   SDDiff = sd(difference), N=length(len))
tg_oj
tg_vc
```

Here we see the two groups, one that received vitamin C through asorbic acid (supp = "VC"), and a group that received vitamin C in orange juice (supp = "OJ").  We also see that the average toot length increases with the dose, and finally that average growth with asorbic acid only eceeds orange juice at the 2 mg dose.  

### 3. Comparison of average tooth growth by Supplement and Dose

#### Using mean tooth length as the comparison
```{r}
tga <- summarise(group_by(ToothGrowth, Supplement = supp, Dose = dose), 
                Mean = mean(len), N = length(len), SD = sd(len))

tgt <- summarise(group_by(tga, Supplement, Dose, Mean), 
                 Interval = qnorm(0.95)*SD/sqrt(N))
summarise(group_by(tgt, Supplement, Dose, Mean), 
          LowerConf = round(Mean - Interval, 2) , 
          UpperConf = round(Mean + Interval, 2))
```

The above summay shows average tooth growth grouped by supplement and dose. It also includes the lower and upper boundaries for the 95% confidence interval of that mean.  This boxplot below illustrates this data:  

```{r echo=FALSE}
boxplot(len~supp*dose, 
        data=ToothGrowth, 
        notch=FALSE, 
        col=(c("red", "green")), 
        main="Tooth Growth", 
        xlab="Supplement and Dose")
```

#### Using mean growth difference as the comparison
```{r}
tg_oj_s <- summarise(group_by(tg_oj, Supplement, Dose, MeanDiff), 
                     Interval = qnorm(0.95)*SDDiff/sqrt(N))
summarise(group_by(tg_oj_s, Supplement, Dose, MeanDiff), 
          LowerConf = round(MeanDiff - Interval, 2) , 
          UpperConf = round(MeanDiff + Interval, 2))
```



# 4. Conclusions. 

The basic rule of thumb is that $\frac{1}{\sqrt{n}}$ gives us a margin of error.  Each supplement only have 10 data points for each dose, which gives a vey high margin of error: **`r percent(1/sqrt(10))`**.  I am basing my conclusion on the assumption that the data has a **normal distribution**.  That said it seems that OJ is the most effective supplement for promoting tooth growth, and since the mean at 2 mg for OJ is included in the upper confidence level for 1.0 mg of OJ it may be that **the optimal dose is 1mg of orange** juice, __assuming__ we want to maximize the effect and minimize the dose.

