---
title: "Machine Learning - HAR Project"
includes:
  in_header: mystyles.sty
output:
  html_document: default
  pdf_document:
    fig_caption: yes
bibliography: bibliography.bib
---

# Assignment Background
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing **how you built your model**, how you used **cross validation**, what you think the **expected out of sample error** is, and why you made the **choices** you did. You will also use your prediction model to **predict 20 different test cases**. 

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the write up to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).

2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details. 


```{r, results="hide", echo=FALSE}
# Load libraries
library(dplyr, quietly=TRUE, warn.conflicts = FALSE)
suppressWarnings(library(scales, quietly=TRUE))
suppressWarnings(library(pander, quietly=TRUE))
library(ggplot2)
suppressWarnings(library(gridExtra, quietly=TRUE, warn.conflict=FALSE))
library(lattice)
library(caret)
library(AppliedPredictiveModeling)
```


# Overview 
The data is based on the HAR study [@har2012], where six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (**Class A**), throwing the elbows to the front (**Class B**), lifting the dumbbell only halfway (**Class C**), lowering the dumbbell only halfway (**Class D**) and throwing the hips to the front (**Class E**).  More information is available from the website here: http://groupware.les.inf.puc-rio.br/har

I will use data from accelerometers on the **belt**, **forearm**, **arm**, and **dumbell** of these participants.  The goal of this project is to predict the manner in which they did the exercise.  This is the **"classe"** variable in the training set.

## Data aquisition and cleanup
```{r cache=TRUE, echo = TRUE}
# Download the test and training data sets
if (!file.exists("./data")) dir.create("./data")

# Training data
tdf <- "./data/pml-training.csv"
if (!file.exists(tdf))
    download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", 
                  destfile=tdf, method="curl")
# Treat the division by zero erros as NA
traindat <- read.csv(tdf, na.strings = c("NA", "#DIV/0!", "", "NaN"))

# Testing data
tstf <- "./data/pml-testing.csv"
if (!file.exists(tstf))
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", 
                destfile=tstf, method="curl")
testdat <- read.csv(tstf, na.strings = c("NA", "#DIV/0!", "", "NaN"))
```

I'm treating *division by zero* errors, empty string, "NaN" and "NA", as *"Not Available"* - **NA**.  

```{r}
nrow(traindat)
cc <- complete.cases(traindat)
sum(!cc)
sum(cc)
#nrow(mc)
#sum
traindat <- subset(traindat, new_window == 'yes')
nrow(traindat)
```

```{r}
# Exploration
library(dplyr)
pander(tally(group_by(traindat, classe), sort = TRUE))

# Explore missing values
library(VIM)
#pander(aggr(traindat, plot=FALSE, combined=TRUE, numbers = FALSE, prop = TRUE))
aggr(traindat, plot=FALSE, only.miss=TRUE)

#traindat_original <- traindat
#str(traindat)
```

# Building a Model  

Describe how I built my model, and why I have made the choices I did.

```{r}
set.seed(34542)
# Only use the variables associated with the focus areas: belt, forearm, arm, and dumbell
training <- select(traindat, classe, 
                   contains('_belt'), 
                   contains('_forearm'), 
                   contains('_arm'), 
                   contains('_dumbell'))
# Remove columns that are mostly NA
training <- select(training, 
                   -starts_with('kurtosis_yaw'),
                   -starts_with('skewness_yaw'))
# Remove the max, min, avg, var & stddev summary variables as these are mostly NA
# training <- select(training, 
#                    -starts_with('var_'),
#                    -starts_with('avg_'),
#                    -starts_with('stddev_'),
#                    -starts_with('max_'),
#                    -starts_with('min_'))

# Do we have any more NAs?
aggr(training, plot=FALSE, only.miss=TRUE)
histMiss(training)
summary(training)
```

## Build a model
```{r}
#str(training)
#summary(training)
#training.imputed <- impute.knn(as.matrix(training))
#preProc <- preProcess(as.matrix(training), method="knnImpute")
# modelFit <- train(classe~., data=training, 
# #                   na.action = na.omit,
# #                 preProcess=c("knnImpute"), # Use K-Nearest-Neighbor to impute values for all the NAs
#                  method="rf") # use random forest 
# modelFit
```

# Cross Validataion

# Receiver Operating Caracteristic (ROC) 


# Accuracy

Expected out of sample error.

# Predictions

Predict 20 different test cases.

# Bibliography
