---
title: "Capstone Notes"
author: "Nils"
date: "18. desember 2015"
output:
  html_document: default
  pdf_document:
    fig_caption: yes
includes:
  in_header: mystyles.sty
---

```{r, results="hide", echo=FALSE}
# Load libraries
library(dplyr, quietly=TRUE, warn.conflicts = FALSE)
suppressWarnings(library(scales, quietly=TRUE))
suppressWarnings(library(pander, quietly=TRUE))
library(ggplot2)
suppressWarnings(library(gridExtra, quietly=TRUE, warn.conflict=FALSE))


library(tm)
library(stringr)
library(SnowballC)
library(tau)
library(class)
library(data.table)
library(RWeka)
```

```{r load source code, message=FALSE, warning=FALSE}
source("./TMWordList.R")
```


# Task 1: Data acquisition and cleaning

## 1.11. Tokenization 

   - Identify appropriate tokens such as words, punctuation, and numbers. 

**Note**:  I will use the TM package for this

   - Write a function that takes a file as input and returns a tokenized version of it.
   
```{r}
readCorpus
```
   
## 1.2. Profanity filtering 

   - removing profanity and other words you do not want to predict.

**Note**: I have a custom profanity dictionary that I use for word removal.  This is applied though the removeWords TM transformation.

```{r}
applyTransformations
```

### Sample corpus 

```{r}
# Sample document (Twitter extract)
readFileLines("./extracts/en_US.twitter.extract.txt")
```


```{r}
# Read sample corpus
#source.dir <- "./Coursera-SwiftKey/final/en_US/"
source.dir <- "./extracts"
language <- "english"
docs.en.sample <- applyTransformations(corpus = readCorpus(source.dir, language),
                                       language = language)
```

```{r}
# Sample corpus after the transformations
docs.en.sample[[1]]$content
```


# Task 2: Exploratory Analysis

```{r Reading full Engligh Corpus, cache=T}
# Read full English corpus
source.dir <- "./Coursera-SwiftKey/extract/en_US/"
language <- "english"
docs.en <- applyTransformations(corpus = readCorpus(source.dir, language),
                                language = language)
```

```{r}
summary(docs.en)
```

## 2.1. Exploratory analysis 

- perform a thorough exploratory analysis of the data, understanding the distribution of words and relationship between the words in the corpora. 

### Count-based evaluation

```{r Create Term Doc Matrix, cache=TRUE }
# Get Term Doc Matrix 
docs.en.tdm <- TermDocumentMatrix(docs.en, control=list(stopwords = TRUE))
# Get words that occur at least 100 times in each doc
f100 <- findFreqTerms(docs.en.tdm, 100)
inspect(docs.en.tdm[f100, 1:3])
```


# 2.2. Understand frequencies of words and word pairs 

- build figures and tables to understand variation in the frequencies of words and word pairs in the data. 

```{r Create Bigram Matrix, cache=TRUE }
# Get Bigram Term Doc Matrix 
bigram.tdm <- getBigramTDM(docs.en)

bigram.dt <- convertTDMtoDataTable(bigram.tdm)

# Get words that occur at least 100 times in each doc
#f100 <- findFreqTerms(docs.en.tdm, 100)
#inspect(docs.en.tdm[f100, 1:3])
```


**Questions to consider:**

1. Some words are more frequent than others - what are the distributions of word frequencies? 

2. What are the frequencies of 2-grams and 3-grams in the dataset? 

3. How many unique words do you need in a frequency sorted dictionary to cover 50% of all word instances in the language? 90%? 

4. How do you evaluate how many of the words come from foreign languages? 

5. Can you think of a way to increase the coverage?

    -- identifying words that may not be in the corpora or using a smaller number of words in the dictionary to cover the same number of phrases?



# Task 3: Modeling

**Questions to consider**

1. How can you efficiently store an n-gram model (think Markov Chains)? 

2. How can you use the knowledge about word frequencies to make your model smaller and more efficient?

3. How many parameters do you need (i.e. how big is n in your n-gram model)?

4. Can you think of simple ways to "smooth" the probabilities 
    (think about giving all n-grams a non-zero probability even if they aren't observed in the data) ?

5. How do you evaluate whether your model is any good? 

6. How can you use backoff models to estimate the probability of unobserved n-grams?
                                                              
    1. Think about starting simply, using your sampled data 

        - What would be your first choice word prediction if you only knew the frequency of each individual word? 

    2. How would you build the model if you were predicting using only one previous word? 



## 3.1. Build basic n-gram model 

- using the exploratory analysis you performed, build a basic [n-gram model](http://en.wikipedia.org/wiki/N-gra) for predicting the next word based on the previous 1, 2, or 3 words.

## 3.2. Build a model to handle unseen n-grams 

- in some cases people will want to type a combination of words that does not appear in the corpora. 

- Build a model to handle cases where a particular n-gram isn't observed. 


# Task 4: Prediction

The goal of this exercise is to build and evaluate your first predictive model. You will use the n-gram and backoff models you built in previous tasks to build and evaluate your predictive model. The goal is to make the model efficient and accurate. 

**Tasks to accomplish**

1.	Build a predictive model based on the previous data modeling steps - you may combine the models in any way you think is appropriate. 

2.	Evaluate the model for efficiency and accuracy - use timing software to evaluate the computational complexity of your model. Evaluate the model accuracy using different metrics like perplexity, accuracy at the first word, second word, and third word. 

**Questions to consider**

1.	How does the model perform for different choices of the parameters and size of the model? 

2.	How much does the model slow down for the performance you gain?

3.	Does perplexity correlate with the other measures of accuracy?

4.	Can you reduce the size of the model (number of parameters) without reducing performance? 

**Tips and hints**

1.	How does the model that just predicts the most frequent words work in terms of accuracy?

2.	How much improvement do you get by using just the previous word to predict the next? 

3.	Consider building the models on smaller samples to check how they perform before scaling up to the whole data set. 

4.	When you fit a predictive model at this step, check the size of the object using the object.size() function, try to build models that have the smallest size possible while retaining accuracy. 

