---
title: "Regression Models - Course Project"
author: "Nils Sandøy"
date: "September 21, 2015"
output: pdf_document
includes:
  in_header: mystyles.sty
---

```{r, results="hide", echo=FALSE}
# Load libraries
library(dplyr, quietly=TRUE, warn.conflicts = FALSE)
suppressWarnings(library(scales, quietly=TRUE))
suppressWarnings(library(pander, quietly=TRUE))
library(ggplot2)
suppressWarnings(library(gridExtra, quietly=TRUE, warn.conflict=FALSE))

library(datasets)
```


**Using the mtcars data set, answer the following two questions:**

* "Is an automatic or manual transmission better for MPG"
* "Quantify the MPG difference between automatic and manual transmissions"

**Grading criteria:**

* Did the student interpret the coefficients correctly?
* Did the student do some exploratory data analyses?
* Did the student fit multiple models and detail their strategy for model selection?
* Did the student answer the questions of interest or detail why the question(s) is (are) not answerable?
* Did the student do a residual plot and some diagnostics?
* Did the student quantify the uncertainty in their conclusions and/or perform an inference correctly?
* Was the report brief (about 2 pages long) for the main body of the report and no longer than 5 with supporting appendix of figures?
* Did the report include an executive summary?
* Was the report done in Rmd (knitr)?


# Executive Summary

# Exploratory Analysis
```{r echo = FALSE}
data(mtcars)

mtcars$am = as.character(mtcars$am)
automatic <- subset(mtcars, am == 1)
manual <- subset(mtcars, am == 0)
```

The mtcars data set consists of `r nrow(mtcars)` rows, `r nrow(automatic)` with automatic and  `r nrow(manual)` with manual transmission.  Some key variables are:

* **mpg** = Miles(US)/gallon
* **am** = Transmission (0 = automatic, 1 = manual)
* **hp** = Gross horsepower 
* **wt** = Weight (lb/1000) 

```{r echo = FALSE}
# Combined scatterplot for the mtcars data, hp vs mpg
g = ggplot(mtcars, aes(x=hp, y=mpg, colour = am, size = wt)) +
    geom_point(alpha=.5) + 
    geom_smooth(method = "lm", size = 2) +
    geom_vline(xintercept = mean(automatic$hp), color = "cyan", alpha=.5) +
    geom_hline(yintercept = mean(automatic$mpg), color = "cyan", alpha=.5) +
    geom_vline(xintercept = mean(manual$hp), color = "salmon", alpha=.5) +
    geom_hline(yintercept = mean(manual$mpg), color = "salmon", alpha=.5) +
    xlab("HP") + 
    ylab("Miles per gallon") + 
    ggtitle("Automatic vs Manual")
g
```

The above scatter plot shows automatic cars in cyan and manual in salmon.  The size of the dots indicate weight. Lighter automatic cars seem to cluster in the upper left quadrant, and are more fuel efficient than heavy manual cars which seem overly represented in the lower right quadrant.  

```{r echo = FALSE}
afit <- lm(mpg ~ I(hp - mean(hp)), automatic)
acoef <- summary(afit)$coef
panderOptions('table.split.table', Inf) 
pander(acoef, style="rmarkdown", caption="LM Coefficients for Automatic (centered at mean hp)")

mfit <- lm(mpg ~ I(hp - mean(hp)), manual)
mcoef <- summary(mfit)$coef
panderOptions('table.split.table', Inf) 
pander(mcoef, style="rmarkdown", caption="LM Coefficients for Manual (centered at mean hp)")
```

An *automatic* car of average engine power, `r round(mean(automatic$hp), 0)` hp, runs `r round(mean(automatic$mpg), 2)` mpg, compared to *manual* cars of average engine power, `r round(mean(manual$hp), 0)`hp, runs `r round(mean(manual$mpg), 2)` mpg.  Manual cars do tend to have larger engines which do consume more fuel, but adding an additional HP of engine power to an automatic car will result in `r round(acoef[2,1], 4)` change in MPG, which is very similar to manual cars (`r round(mcoef[2,1], 4)`), assuming we ignore weight in this data set. 

## MPG difference between autumatic and manual transmission
| Transmission type | MPG difference per added HP |
|:-----------------:|:---------------------------:|
| Automatic         | `r round(acoef[2,1], 4)`    |
| Manual            | `r round(mcoef[2,1], 4)`    |

From the slope alone there is not much difference in fuel efficiency between the transmission types, but from the scatterplot we see that the confidence interval is much wider for automatic than it is for manual cars, this is because of the wider spread in values.  An investigation into the residuals will show the difference here.


# Conclusion


