---
title: "Week 3 - Quizz Notes"
author: "Nils"
date: "27. september 2015"
output:
  html_document: default
  pdf_document:
    fig_caption: yes
includes:
  in_header: mystyles.sty
---


```{r, results="hide", echo=FALSE}
# Load libraries
library(dplyr, quietly=TRUE, warn.conflicts = FALSE)
suppressWarnings(library(scales, quietly=TRUE))
suppressWarnings(library(pander, quietly=TRUE))
library(ggplot2)
suppressWarnings(library(gridExtra, quietly=TRUE, warn.conflict=FALSE))

library(datasets)
library(car)
require(lmtest)
```

# Question 1

Consider the mtcars data set. Fit a model with mpg as the outcome that includes number of cylinders as a factor variable and weight as confounder. Give the adjusted estimate for the expected change in mpg comparing 8 cylinders to 4.


```{r}
data("mtcars")

fit <- lm(mpg ~ factor(cyl) + wt, data = mtcars)
coef <- summary(fit)$coef

panderOptions('table.split.table', Inf) 
pander(coef, style="rmarkdown", caption="mpg ~ factor(cyl) + wt")

```


# Question 2

Consider the mtcars data set. Fit a model with mpg as the outcome that includes number of cylinders as a factor variable and weight as a possible confounding variable. Compare the effect of 8 versus 4 cylinders on mpg for the adjusted and unadjusted by weight models. Here, adjusted means including the weight variable as a term in the regression model and unadjusted means the model without weight included. What can be said about the effect comparing 8 and 4 cylinders after looking at models with and without weight included?.

```{r}
fit2 <- lm(mpg ~ factor(cyl), data = mtcars)
coef2 <- summary(fit2)$coef

panderOptions('table.split.table', Inf) 
pander(coef2, style="rmarkdown", caption="mpg")
pander(coef, style="rmarkdown", caption="mpg ~ factor(cyl) + wt")
```

Cylinder has less of an impact when considering weight

```{r mod_m, echo = FALSE, fig.height = 5, fig.cap="\\label{fig:mod_m} mpg ~ cyl + wt"}

g1 <- ggplot(mtcars, aes(x=factor(cyl), y=mpg, size=wt)) +
        geom_point(alpha=.5) +
        geom_smooth(method = "lm", se = TRUE) +
#        geom_hline(yintercept = mean(automatic$mpg), color = "cyan") +
#        geom_hline(yintercept = mean(manual$mpg), color = "salmon") +
        theme(legend.position="bottom")
g2 <- ggplot(mtcars, aes(x=factor(cyl), y=mpg)) +
        geom_point(alpha=.5) +
        geom_smooth(method = "lm", se = TRUE) +
#        geom_hline(yintercept = mean(automatic$mpg), color = "cyan") +
#        geom_hline(yintercept = mean(manual$mpg), color = "salmon") +
        theme(legend.position="bottom")
g3 <- ggplot(mtcars, aes(x=wt, y=mpg, size=wt)) +
        geom_point(alpha=.5) +
        geom_smooth(method = "lm", se = TRUE) +
#        geom_hline(yintercept = mean(automatic$mpg), color = "cyan") +
#        geom_hline(yintercept = mean(manual$mpg), color = "salmon") +
        theme(legend.position="bottom")
g4 <- ggplot(mtcars, aes(x=wt, color=factor(cyl), y=mpg)) +
        geom_point(alpha=.5) +
        geom_smooth(method = "lm", se = TRUE) +
#        geom_hline(yintercept = mean(automatic$mpg), color = "cyan") +
#        geom_hline(yintercept = mean(manual$mpg), color = "salmon") +
        theme(legend.position="bottom")
grid.arrange(g1, g2, g3, g4, ncol=2, nrow=2)
```


# Question 3
Consider the  mtcars  data set. 

Fit a model with mpg as the outcome that considers number of cylinders as a factor variable and weight as confounder. 
Fit a second model with mpg as the outcome model that considers the interaction between number of cylinders (as a factor variable) and weight. 

Give the P-value for the likelihood ratio test comparing the two models and suggest a model using 0.05 as a type I error rate significance benchmark. 

```{r}
fit1 <- lm(mpg ~ factor(cyl) + wt, data = mtcars)
fit2 <- lm(mpg ~ factor(cyl) * wt, data = mtcars)

panderOptions('table.split.table', Inf) 
pander(summary(fit1)$coef, style="rmarkdown", caption="mpg ~ factor(cyl) + wt")
pander(summary(fit2)$coef, style="rmarkdown", caption="mpg ~ factor(cyl) * wt")

pander(lrtest(fit1, fit2), style="rmarkdown")
```


# Question 4

Consider the mtcars data set. Fit a model with mpg as the outcome that includes number of cylinders as a factor variable and weight inlcuded in the model as

```{r}
fit <- lm(mpg ~ I(wt * 0.5) + factor(cyl), data = mtcars)
pander(summary(fit)$coef, style="rmarkdown")
```

How is the wt coefficient interpretted?

* The estimated expected change in MPG per half ton increase in weight for the average number of cylinders. 
* The estimated expected change in MPG per one ton increase in weight. 
* The estimated expected change in MPG per one ton increase in weight for a specific number of cylinders (4, 6, 8). 
* The estimated expected change in MPG per half ton increase in weight. 
* The estimated expected change in MPG per half ton increase in weight for for a specific number of cylinders (4, 6, 8).



# Question 5
Consider the following data set
```{r}
x <- c(0.586, 0.166, -0.042, -0.614, 11.72)
y <- c(0.549, -0.026, -0.127, -0.751, 1.344)

```

Give the hat diagonal for the most influential point

```{r}
fit <- lm(y ~ x)
hv <- hatvalues(fit)
pander(hv, style="rmarkdown")

max(hv)

```

* 0.2287 
* 0.2025 
* 0.9946 
* 0.2804


# Question 6

Consider the following data set
```{r}
x <- c(0.586, 0.166, -0.042, -0.614, 11.72)
y <- c(0.549, -0.026, -0.127, -0.751, 1.344)
```

Give the slope dfbeta for the point with the highest hat value.

```{r}
fit <- lm(y ~ x)
pander(hatvalues(fit), style="rmarkdown")
pander(dfbetas(fit), style="rmarkdown")
```

* 0.673 
* -134 
* -.00134  
* -0.378 

# Question 7

Consider a regression relationship between Y and X with and without adjustment for a third variable Z. Which of the following is true about comparing the regression coefficient between Y and X with and without adjustment for Z.


* The coefficient can't change sign after adjustment, except for slight numerical pathological cases.  
* For the the coefficient to change sign, there must be a significant interaction term. 
* Adjusting for another variable can only attenuate the coefficient toward zero. It can't materially change sign. 
* It is possible for the coefficient to reverse sign after adjustment. For example, it can be strongly significant and positive before adjustment and strongly significant and negative after adjustment.
