---
title: "RegModels Exercises Week 4"
author: "Nils Sandøy"
date: "September 29, 2015"
output: pdf_document
---


```{r, results="hide", echo=FALSE}
# Load libraries
library(dplyr, quietly=TRUE, warn.conflicts = FALSE)
suppressWarnings(library(scales, quietly=TRUE))
suppressWarnings(library(pander, quietly=TRUE))
library(ggplot2)
suppressWarnings(library(gridExtra, quietly=TRUE, warn.conflict=FALSE))

library(datasets)
library(car)
```


1. Load the dataset Seatbelts as part of the datasets package via data(Seatbelts). Use as.data.frame to convert the object to a dataframe. Create a new outcome variable for whether or not greater than 119 drivers were killed that month. Fit a logistic regression GLM with this variable as the outcome and kms, PetrolPrice and law as predictors. Interpret your parameters. [Watch a video solution](https://www.youtube.com/watch?v=CXWZqzKdkp4&list=PLpl-gQkQivXji7JK1OP1qS7zalwUBPrX0&index=52).


```{r}
data("Seatbelts")
# Create a new outcome variable for whether or not greater than 119 drivers were killed that month.
sb <- mutate(as.data.frame(Seatbelts),
             dkb <- 1 * (DriversKilled > 119),  # Multiply by 1 forces this to be numeric, rather than boolean
             pp = (PetrolPrice - mean(PetrolPrice)) / sd(PetrolPrice),
             mm = kms / 1000,
             mmc = mm - mean(mm))

# Fit a logistic regression GLM with this variable as the outcome and kms, PetrolPrice and law as predictors
glm <- glm(dkb ~ pp + mmc + law  , data = sb,family="binomial")
coefs <- summary(glm)$coef
panderOptions('table.split.table', Inf) 
pander(coefs, style="rmarkdown", caption="dkb ~ kms + PetrolPrice + law")
```

#### Interpret your parameters.
Use exponentials to interpret the numbers.
```{r}
# Eponentiation of the law coefficient
1 - exp(coefs[4,1])
```

There was a `r percent(1 - exp(coefs[4,1]))` decrease in  the odds of deaths beeing > 119 after the law was enacted, holding the other coefficients fixed.

```{r}
# Eponentiation of the miles coefficient
1 - exp(coefs[3,1])
```

There was a `r percent(1 - exp(coefs[3,1]))` decrease in the odds of deaths beeing > 119 per additional 1000 driver miles per month, holding the other coefficients fixed.


2. Fit a binomial model with DriversKilled as the outcome and drivers as the total count with kms , PetrolPrice and law as predictors, interpret your results. [Watch a video solution](https://www.youtube.com/watch?v=M2KLD_ZFgdo&list=PLpl-gQkQivXji7JK1OP1qS7zalwUBPrX0&index=53). 

```{r}
# Fit a binomial model with DriversKilled as the outcome and drivers as the total count with kms , PetrolPrice and law as predictors
glm <- glm(cbind(DriversKilled, drivers - DriversKilled) ~ pp + mmc + law, data = sb, family="binomial")
coefs <- summary(glm)$coef
panderOptions('table.split.table', Inf) 
pander(coefs, style="rmarkdown", caption="DriversKilled ~ kms + pp + law")
```

mmc & law are now positive, so there is an increase in the odds of deaths

3. Refer to Question 1. Use the anova function to compare models with just law, law and PetrolPrice and all three predictors. [Watch a video solution](https://www.youtube.com/watch?v=npHpBLqkhLg&list=PLpl-gQkQivXji7JK1OP1qS7zalwUBPrX0&index=54). 

```{r}
#  Use the anova function to compare models with just law, law and PetrolPrice and all three predictors
m1 <- glm <- glm(dkb ~ law, data = sb,family="binomial")
m2 <- glm <- glm(dkb ~ law + pp, data = sb,family="binomial")
m3 <- glm <- glm(dkb ~ law + pp + mmc  , data = sb,family="binomial")

panderOptions('table.split.table', Inf) 
pander(anova(m1, m2, m3), style="rmarkdown")

pander(summary(m1)$coef, style="rmarkdown", caption= "m1 summary coefs")
pander(summary(m2)$coef, style="rmarkdown", caption= "m2 summary coefs")
pander(summary(m3)$coef, style="rmarkdown", caption= "m3 summary coefs")
```

m3 doesn't seem to add much value to the model.  **Go with m2.**


# Poisson Exercises

#### 1.
Load the dataset Seatbelts as part of the datasets package via data(Seatbelts). Use as.data.frame to convert the object to a dataframe. Fit a Poisson regression GLM with UKDriversKilled as the outcome and kms, PetrolPrice and law as predictors. Interpret your results. [Watch a video solution](https://www.youtube.com/watch?v=TXO-SHOV_j4&list=PLpl-gQkQivXji7JK1OP1qS7zalwUBPrX0&index=55).


```{r}
m <- glm(DriversKilled ~ pp + mmc + law, data=sb, family = poisson)
pander(m)
pander(exp(m$coef))
pander(1 - exp(m$coef))
```





#### 2.
Refer to question 1. Fit a linear model with the log of drivers killed as the outcome. Interpret your results. [Watch a video solution](https://www.youtube.com/watch?v=7RyaIhmpM48&list=PLpl-gQkQivXji7JK1OP1qS7zalwUBPrX0&index=56). 

#### 3.
Refer to question 1. Fit your Poisson log-linear model with drivers as a log offset (to consider the proportion of drivers killed of those killed or seriously injured.) [Watch a video solution](https://www.youtube.com/watch?v=HylRM_XrUe0&list=PLpl-gQkQivXji7JK1OP1qS7zalwUBPrX0&index=57). 

#### 4.
Refer to Question 1. Use the anova function to compare models with just law, law and PetrolPrice and all three predictors. [Watch a video solution](https://www.youtube.com/watch?v=ewfjP1i8gPs&list=PLpl-gQkQivXji7JK1OP1qS7zalwUBPrX0&index=58). 



